---
title: 'Homework 1'
subtitle: ''
author: 'Daria Dubovskaia'
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: united
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: sentence
---


```{r setup, include=FALSE}
# chunks
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, include=TRUE, 
message=FALSE, warning=FALSE, fig.height=5, fig.align='center')

# libraries
library(tidyverse)
library(kableExtra)
library(summarytools)
library(MASS) # glm.nb()
library(mice)
library(lubridate)
library(splusTimeDate)
library(pscl) # zeroinfl()
library(skimr)
library(sjPlot)
library(mpath)
library(yardstick)
library(labelled)
library(haven)
library(corrplot)
library(Hmisc)
library(jtools)
library(caret)
library(broom)

 # library
library(treemap)
library(ggplot2)
library(hrbrthemes)
library(viridis)
library(vip)

library(ggpubr)
library(rockchalk)



# ggplot
theme_set(theme_light())


# random seed
set.seed(42)
```



# Overview

------------------------------------------------------------------------

## 1. Data Exploration and Preparation

The **small** dataset contained `r nrow(df_1000)` observations of 17 predictor variables. 
The **large** dataset contained `r nrow(df_50000)` observations over the same predictor variables.
Each record in both datasets describes a single sales transaction in full, including the product type, sales channel, financial data, and pertinent dates. Specifically:

-    `Region`: Geographical region where the sale occurred
-    `Country`: Country where the sale occurred
-    `Item Type`: Type of product sold (e.g. Cereal, Cosmetics, Fruits, etc.)
-    `Sales Channel`: How the sale was made (Offline/Online)
-    `Order Priority`: Priority of the order (e.g., M for medium, C for critical, H for high, etc.)
-    `Order Date`: Date when the order was placed
-    `Order ID`: Unique identifier for the order
-    `Ship Date`: Date when the product was shipped
-    `Units Sold`: Number of units sold
-    `Unit Price`: Price per unit of the product
-    `Unit Cost`: Cost per unit of the product
-    `Total Revenue`: Total revenue from the sale
-    `Total Cost`: Total cost of the sale
-    `Total Profit`: Total profit from the sale


```{r data_load}
# load data
df_1000 <- read.csv("C:/Users/daria/Downloads/1000 Sales Records.csv")
df_50000 <- read.csv("C:/Users/daria/Downloads/50000 Sales Records.csv")
``` 



### 1.1 Summary Statistics
```{r head_data_small}
DT::datatable(
      df_1000[1:25,],
      extensions = c('Scroller'),
      options = list(scrollY = 350,
                     scrollX = 500,
                     deferRender = TRUE,
                     scroller = TRUE,
                     dom = 'lBfrtip',
                     fixedColumns = TRUE, 
                     searching = FALSE), 
      rownames = FALSE) 
```

```{r head_data_large}
DT::datatable(
      df_50000[1:25,],
      extensions = c('Scroller'),
      options = list(scrollY = 350,
                     scrollX = 500,
                     deferRender = TRUE,
                     scroller = TRUE,
                     dom = 'lBfrtip',
                     fixedColumns = TRUE, 
                     searching = FALSE), 
      rownames = FALSE) 
```


```{r summary_small}
DT::datatable(
      dfSummary(df_1000, text.graph.col = FALSE, graph.col = FALSE),
      extensions = c('Scroller'),
      options = list(scrollY = 350,
                     scrollX = 500,
                     deferRender = TRUE,
                     scroller = TRUE,
                     dom = 'lBfrtip',
                     fixedColumns = TRUE, 
                     searching = FALSE), 
      rownames = FALSE) 

```

```{r summary_large}
DT::datatable(
      dfSummary(df_50000, text.graph.col = FALSE, graph.col = FALSE),
      extensions = c('Scroller'),
      options = list(scrollY = 350,
                     scrollX = 500,
                     deferRender = TRUE,
                     scroller = TRUE,
                     dom = 'lBfrtip',
                     fixedColumns = TRUE, 
                     searching = FALSE), 
      rownames = FALSE) 

```

### 1.2 Column change, New columns


Change column names to lower case, substitute " " with "_".
```{rcol_name_function}
 
# This function changes the column names of a data frame to lower case and substitutes dots with underscores
#
# PARAMETERS:
#   - df: The input data frame
#     TYPE: data frame
#
# RETURNS:
#   - The modified data frame with column names substituted
#     TYPE: data frame
#
# EXCEPTIONS:
#   - This function assumes that the input is a valid data frame and does not handle other types of inputs

change_column_names <- function(df) {
  # Get the current column names of the data frame
  
  colnames(df) <- colnames(df)  %>%
  str_to_lower()  %>%
  str_replace_all("\\.", "_")
  
  # Return the modified data frame
  return(df)
}
```

```{r col_name}
# Change column names, lower case, substitute "." with "_"
df_1000 <- change_column_names(df_1000)
df_50000 <- change_column_names(df_50000)
```

Change columns Order Date and Ship Date to date format

```{r to_date_function}

# This function transforms a character column from a dataset in the format "dd/mm/yyyy" to a date format
#
# PARAMETERS:
#   - df: A data frame containing the dataset
#     TYPE: data frame
#
# RETURNS:
#   - The input data frame with the specified column transformed to a date format
#     TYPE: data frame
#
# EXCEPTIONS:
#   - This function assumes that the character column contains valid dates in the format "dd/mm/yyyy"
#     It does not handle missing values or invalid date formats


	transform_date_columns <- function(df) {
  # Get the column names of the dataframe
  column_names <- colnames(df)
  
  # Iterate over each column in the dataframe
  for (column_name in column_names) {
    # Check if the column is of character type
    if (is.character(df[[column_name]])) {
      # Check if the column values are in the format dd/mm/yyyy
      if (any(grepl("^\\d{2}/\\d{2}/\\d{4}$", df[[column_name]]))) {
        # Transform the column values to the date format, %Y to show years 2000–2068 properly
        df[[column_name]] <- as.Date(df[[column_name]], format="%m/%d/%Y")
      }
    }
  }
  
  # Return the transformed dataframe
  return(df)
}
```

```{r to_date}
# Transform the character column to a date format
df_1000 <- transform_date_columns(df_1000)
df_50000 <- transform_date_columns(df_50000)
```

New columns factor Quarter, dummy Weekend, dummy for sales channel

```{r col_quarter_function}
#   This function adds a new column called "quarter" to a data frame based on the values in a date column
#   The "quarter" column represents the quarter of the year corresponding to each date
#
# PARAMETERS:
#   - df: The input data frame
#     TYPE: data frame
#
#   - date_column: The name of the date column in the input data frame
#     TYPE: date
#
# RETURNS:
#   - The input data frame with an additional "quarter" column
#     TYPE: data frame
#
# EXCEPTIONS:
#   - This function assumes that the input data frame contains a valid date column
#   - If the date column is not found in the input data frame, an error will be thrown
#

	add_quarter_column <- function(df, date_column) {
  # Check if the date column exists in the input data frame
  if (!(date_column %in% colnames(df))) {
    stop("Date column not found in the input data frame.")
  }
  
  # Extract the quarter from the date column using lubridate::quarter function
  df$quarter <- factor(quarter(df[[date_column]]))
  
  # Return the modified data frame with the additional "quarter" column
  return(df)
}
```

```{r col_weekend_function}
#   This function takes a dataframe as input and adds a new column indicating whether the day of the order date
#   is a weekend or not. If the day is a weekend, the value in the new column will be 1, otherwise it will be 0
#
# PARAMETERS:
#   - df: The input dataframe containing the order date column
#     TYPE: dataframe
#
#   - date_col: The name of the column in the dataframe that contains the order dates
#     TYPE: date
#
# RETURNS:
#   - The input dataframe with an additional column indicating whether the day is a weekend or not
#     TYPE: dataframe
#
# EXCEPTIONS:
#   - This function assumes that the order date column is in the correct format (YYYY-MM-DD) and does not handle
#     invalid date formats.
#   - This function assumes that the order date column contains valid dates and does not handle missing or null values

 
add_weekend_indicator <- function(df, date_col) {

  # Create a new column to store the weekend indicator
  df$is_weekend <- factor(ifelse(weekdays(df[[date_col]]) %in% c("Saturday", "Sunday"), 1, 0))
  
  # Return the modified dataframe
  return(df)
}

```

```{r col_online_function}
# This function creates a dummy variable based on the values of a specified column in a data frame
# If the sales channel is a online, the value in the new column will be 1, otherwise it will be 0
#
# PARAMETERS:
#   - data: The data frame containing the column to create the dummy variable from
#     TYPE: data frame
#
#   - channel_column: The name of the column in the data frame to create the dummy variable from
#     TYPE: character
#
# RETURNS:
#   - The modified data frame with the dummy variable added as a new column
#     TYPE: data frame

#
# EXCEPTIONS:
#   - This function assumes that the specified column exists in the data frame
#   - This function assumes that the specified column contains values of either "Online" or "Offline"

add_online_indicator <- function(df, channel_column) {
  # Check if the specified column exists in the data frame
  if (!(channel_column %in% colnames(df))) {
    stop("Specified column does not exist in the data frame.")
  }
  
  # Create the dummy variable 'is_online'
  df$is_online <- factor(ifelse(df[[channel_column]] == "Online", 1, 0))
  
  # Return the modified data frame
  return(df)
}
```


```{r new_columns}
# Create new columns quarter, is_weekend, is_online
df_1000 <- df_1000 %>%
          add_quarter_column("order_date") %>%
          add_weekend_indicator("order_date") %>%
          add_online_indicator("sales_channel")

df_50000 <- df_50000 %>%
          add_quarter_column("order_date") %>%
          add_weekend_indicator("order_date") %>%
          add_online_indicator("sales_channel")
```



Character columns to factor 

```{r to_factor_function}
# This function takes a data frame as input and transforms all character columns to factor columns
#
# PARAMETERS:
#   - df: A data frame
#     TYPE: data.frame
#
# RETURNS:
#   - The transformed data frame with character columns converted to factor columns
#     TYPE: data.frame
#
# EXCEPTIONS:
#   - This function assumes that all character columns in the data frame can be safely converted to factors
#   - If a column contains non-character values, an error may occur

transform_character_columns <- function(df) {
  # Identify the character columns in the data frame
  char_cols <- sapply(df, is.character)
  
  # Transform the character columns to factor columns
  df[char_cols] <- lapply(df[char_cols], as.factor)
  
  # Return the transformed data frame
  return(df)
}
```

```{r to_factor}
# Transform to factor character columns
df_1000 <- transform_character_columns(df_1000)
df_50000 <- transform_character_columns(df_50000)
```


Order ID to character

```{r to_character_function}
# This function takes a data frame as input and transforms a numeric column to a character column
#
# PARAMETERS:
#   - df: A data frame
#     TYPE: data.frame
#
# RETURNS:
#   - The transformed data frame with a numeric column converted to a character column
#     TYPE: data.frame
#
# EXCEPTIONS:
#   - This function assumes that any numeric column in the data frame can be safely converted to characters
#   - If a column contains non-numeric values, an error may occur

transform_numeric_column <- function(df, num_column) {
  
  # Transform the numeric column to a character column
  df[num_column] <- lapply(df[num_column], as.character)
  
  # Return the transformed data frame
  return(df)
}
```

```{r to_character}
# Transform from numeric to character order_id column
df_1000 <- transform_numeric_column(df_1000, "order_id")
df_50000 <- transform_numeric_column(df_50000, "order_id")
```

Summary for the updated data


```{r summary}
DT::datatable(
      dfSummary(df_1000, text.graph.col = FALSE, graph.col = FALSE),
      extensions = c('Scroller'),
      options = list(scrollY = 350,
                     scrollX = 500,
                     deferRender = TRUE,
                     scroller = TRUE,
                     dom = 'lBfrtip',
                     fixedColumns = TRUE, 
                     searching = FALSE), 
      rownames = FALSE) 

```





